\documentclass[11pt, preprint]{article}
\usepackage{hyperref}
\usepackage{rotating}
 
\setlength{\footnotesep}{9.6pt}
\setlength{\parskip}{12pt}

\newcounter{thefigs}
\newcommand{\fignum}{\arabic{thefigs}}

\newcounter{thetabs}
\newcommand{\tabnum}{\arabic{thetabs}}

\newcounter{address}

\begin{document}

\begin{center}
  {\bf Radiative Processes in Astrophysics / Problem Set \#1 \\
    Due February 15, 2021}
\end{center}

\noindent Pick {\bf two} of the problems below to answer. Submit
as a PDF by email!

\begin{enumerate}
\item Show that for an optically thin cloud around a source with mass
  $M$ and luminosity $L$ (integrated over frequency), the condition
  that the luminosity drives away the cloud through radiation pressure
  is $L>4\pi G c/\kappa$, where $\kappa$ is the opacity
  (i.e. absorption coefficient per unit mass, assumed constant with
  frequency). 
\item Sunspots on the Sun have a temperature of $\sim 4000$ K,
  relative to the typical location on the Sun, which has a temperature
  of $\sim 5500 K$. What is the ratio of the specific intensity
  integrated over frequency ($\int d\nu I_\nu$) that we should observe
  in the location of a Sunspot relative to a typical location on the
  solar disk?
\item Light that reaches us from the Sun's ``limb'' (the area near its
  apparent angular edge) has a lower specific intensity than light
  that reaches us from its angular center. Explain why. Do you predict
  that the spectrum as a function of frequency differs in shape from
  center to edge, and if so how?
\item Given that for redshifted light $v_o = \nu_e / (1+z)$ show that
  $I_\nu \nu^{-3}$ is a constant. 
\item In cosmology the angular diameter distance is defined as $D_A =
  s / \theta$, where $s$ is the physical size of an object and
  $\theta$ is its angular size, and the luminosity distance is defined
  as $D_L = \sqrt{L/4\pi f}$, where $L$ is the total luminosity of an
  object (integrated over all frequencies) and $f$ is its measured
  flux. Using only the fact that $I_\nu \nu^{-3}$ is a constant show
  that $D_L/D_A = (1+z)^2$.
\end{enumerate}

\end{document}

